Local PCA Shows How Population Structure Differs Along the Genome

### 2016-05-26 writeup/local_pca_paper.tex
added writeup of weighted pca to appendix

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig1}
    \end{center}
    \caption{
         Variation in structure for windows on chromosome arms 2L, 2R, 3L, 3R and X. 
         Each represents a window: the first column shows the MDS visualization of relationships between windows. 
         The second and third columns show the position (midpoint) of each window against the first and second MDS coordinates, respectively. 
         Rows show results from different chromosome arms, from top to bottom: Chr2L, Chr2R, Chr3L, Chr3R, ChrX. 
         Colors are consistent for plots in each row. 
         Vertical black lines show the breakpoints of known polymorphic inversions for Chr2L, Chr2R, Chr3L, Chr3R (from top to down: In(2L)t, In(2R)NS, In(3L)OK, In(3R)K).
         \label{fig:mds_chr2L}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig2}
    \end{center}
    \caption{      
        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L}. 
        Each point represents a sample. 
        The plots are plotting the second principal components against the first principal components. 
        The first column shows the windows colored green in Figure 1; the second column shows orange windows; and third shows purple windows.
         \label{fig:pca_chr2L}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.7\textwidth]{fig5}
    \end{center}
    \caption{
         MDS results and gene density for the \textit{Medicago} genome (chromosome 1-8). 
         For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
         The black plot below is gene count for each window against the position of each window.
         \label{fig:mds_medicago}
    }
\end{figure}

@@

PCA can be thought of as finding a good low-dimensional matrix factorization \citep{engelhardt2010analysis}
that well-approximates the original data in the least-squares sense:
if $C$ is the $N \times N$ matrix of genotypes, 
then to find the top $k$ principal components, 
we find an orthogonal $N \times k$ matrix $U$,
and a $k \times k$ diagonal matrix $\Lambda$ with diagonal entries $\Lambda_{ii}=\lambda_i$ to minimize
\begin{align} \label{eqn:objective}
    \| C - U \Lambda U^T \|^2 = \sum_{ij} \left( G_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 .
\end{align}
The columns of $U$, known as the principal components, are the eigenvectors of $C$,
the eigenvalues of $C$ are $\lambda$, and the proportion of variance explained by the $m^\text{th}$ component is
\begin{align*}
    \frac{ \lambda_m }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m U_{im} U_{jm} )^2 }{ \sum_{ij} C_{ij}^2 } .
\end{align*}

@@

To solve \eqref{eqn:weighted_objective},
let $\lambda$ and $V$ denote the eigenvalues and eigenvectors of $W^{1/2} C W^{1/2}$,
so that $V \Lambda V^T$ is the closest in least squares to $W^{1/2} C W^{1/2}$;
so if we define $U = W^{-1/2} V$ then 
then $U^T W U = V^T V = I$,
and 
\begin{align*}
    W^{-1/2} V \Lambda V^T W^{-1/2} 
    =
    U \Lambda U^T
\end{align*}
is the low-dimensional approximation to $C$.
The proportion of variance explained is calculated exactly as before,
but the variance is computed in $\ell_2(w)$:
\begin{align*}
    \frac{ \lambda_m }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m w_i w_j U_{im} U_{jm} )^2 }{ \sum_{ij} w_i w_j C_{ij}^2 } .
\end{align*}

>>>>

\begin{figure}
    \begin{center}
       \includegraphics[width=1\textwidth]{Fig1_Together_MDS_plot}
    \end{center}
    \caption{
         Variation in structure for windows on chromosome arms 2L, 2R, 3L, 3R and X. 
         Each represents a window: the first column shows the MDS visualization of relationships between windows. 
         The second and third columns show the position (midpoint) of each window against the first and second MDS coordinates, respectively. 
         Rows show results from different chromosome arms, from top to bottom: Chr2L, Chr2R, Chr3L, Chr3R, ChrX. 
         Colors are consistent for plots in each row. 
         Vertical black lines show the breakpoints of known polymorphic inversions for Chr2L, Chr2R, Chr3L, Chr3R (from top to down: In(2L)t, In(2R)NS, In(3L)OK, In(3R)K).
         \label{fig:mds_chr2L}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.9\textwidth]{Fig2_all_pca_plots}
    \end{center}
    \caption{      
        PCA plots for the three sets of genomic windows colored in Figure \ref{fig:mds_chr2L}. 
        Each point represents a sample. 
        The plots are plotting the second principal components against the first principal components. 
        The first column shows the windows colored green in Figure 1; the second column shows orange windows; and third shows purple windows.
         \label{fig:pca_chr2L}
    }
\end{figure}

@@

\begin{figure}
    \begin{center}
       \includegraphics[width=0.7\textwidth]{Medicago_MDS_1D_and_gene_count}
    \end{center}
    \caption{
         MDS results and gene density for the \textit{Medicago} genome (chromosome 1-8). 
         For each chromosome, the red plot above is first coordinate of MDS against the middle position of each window along each chromosome. 
         The black plot below is gene count for each window against the position of each window.
         \label{fig:mds_medicago}
    }
\end{figure}

@@

PCA can be thought of as finding a good low-dimensional matrix factorization \citep{engelhardt2010analysis}
that well-approximates the original data in the least-squares sense:
if $C$ is the $N \times N$ matrix of genotypes, 
then to find the top $k$ principal components, 
we find an orthogonal $N \times k$ matrix $U$,
and a $k \times k$ diagonal matrix $\Lambda$ with diagonal entries $\Lambda_{ii}=\lambda_i$ to minimize
\begin{align} \label{eqn:objective}
    \| C - U \Lambda U^T \|^2 = \sum_{ij} \left( G_{ij} - \sum_m \lambda_{m} U_{im} U_{jm} \right)^2 .
\end{align}
The columns of $U$, known as the principal components, are the eigenvectors of $C$,
the eigenvalues of $C$ are $\lambda$, and the proportion of variance explained by the $m^\text{th}$ component is
\begin{align*}
    \frac{ \lambda_m^2 }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m U_{im} U_{jm} )^2 }{ \sum_{ij} C_{ij}^2 } .
\end{align*}

@@

To solve \eqref{eqn:weighted_objective},
let $\lambda$ and $V$ denote the eigenvalues and eigenvectors of $W^{1/2} C W^{1/2}$,
so that $V \Lambda V^T$ is the closest in least squares to $W^{1/2} C W^{1/2}$;
so if we define $U = W^{-1/2} V$ then 
then $U^T W U = V^T V = I$,
and 
\begin{align*}
    W^{-1/2} V \Lambda V^T W^{-1/2} 
    =
    U \Lambda U^T
\end{align*}
is the low-dimensional approximation to $C$.
The proportion of variance explained is calculated exactly as before,
but the variance is computed in $\ell_2(\sqrt{w})$:
\begin{align*}
    \frac{ \lambda_m^2 }{ \sum_\ell \lambda_\ell^2 } = \frac{ \sum_{ij} ( \lambda_m \sqrt{w_i w_j} U_{im} U_{jm} )^2 }{ \sum_{ij} \sqrt{w_i w_j} C_{ij}^2 } .
\end{align*}
